Intrusion Detection with \\ Genetic Algorithms and Fuzzy Logic

### 2013-12-06 Talk/Presentation.tex
Added another slide on encoding of features and rules, added to the mutation slide, added stuff to the fitness function slide, added more to the conclusions side.

@@

\begin{frame}
  \frametitle{RLD09}
	\begin{itemize}
		\item RLD09: created because KDD99 is 14 years old,

@@

\begin{frame}
  \frametitle{Rules}
	\begin{itemize}
		\item A commonly used approach for detecting intrusions is to use rules.
		\item If-Then format: If (\emph{condition}) then (\emph{consequence}).
		\begin{itemize}
			\item Condition: one or more features
			\item Consequence: says if it is an intrusion or not.
			\linespace
			\item If \emph{duration = 4} then \emph{intrusion}.
		\end{itemize}				
	\end{itemize}
\end{frame}

@@

\subsection{Determining the Accuracy of an Algorithm}
\begin{frame}
  \frametitle{Determining the Accuracy of an Algorithm}
\begin{table}
Predicted
\begin{tabular}{l|ll}
Actual   & Not Attack & Attack \\ \hline
Not Attack & True Negative (TN) & False Positive (FP) \\
Attack & False Negative (FN) & True Positive (TP)  \\
\end{tabular}
\end{table}
\linespace
\linespace
\begin{center}
Detection rate (DR): percentage of normal and attack activity correctly classified from the total number of data records.
\end{center}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Fuzzy Algorithm]{Fuzzy Algorithm}
\subsection{Fuzzy Logic}
\begin{frame}
	\frametitle{Fuzzy Logic}
	\begin{itemize}
		\item Fuzzy logic: used in intrusion detection systems to find degree of certainty of record being attack.
			\begin{itemize}
				\item If it's not clear if activity is an attack, fuzzy logic says where on a spectrum it is and how certain it is of being an attack.
			\end{itemize}
		\linespace
		\linespace
		\linespace
		\item Fuzzy logic rules: similar to rules described before, except that
consequence is certainty factor. 
		\begin{itemize}
			\item If (\emph{duration} = 6.2) then (\emph{the degree of certainty of the record being an attack is 0.8}).
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\begin{figure}
\begin{small}
\begin{tabular}{|cccc|c|cccc|c|} \hline
010 & 011 & 100 & 101   & ... & 010 & 011 & 101 & 111   & Attack\\
a=2 & b=3 & c=4 & d=5   & ... & a=2 & b=3 & c=5 & d=7   &\\ 
    &     & Block 1&    &        &     & Block 12& &       & Type\\
\hline\end{tabular}
\caption{Based on [Jongsuebsuk \emph{et al.}, 2013]}
\end{small}
\end{figure}

	\begin{itemize}
		\item Degree of certainty is computed for each of 12 blocks, if sum of those is greater than threshold, declared as attack.
	\end{itemize}

\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Genetic Algorithms]{Genetic Algorithms}
\subsection{GA Overview}
\begin{frame}
  \frametitle{Genetic Algorithms}
	\begin{itemize}
		\item GAs: search technique used to find solutions to problems.
        \item Possible solutions to problems: represented in a variety of problem dependent ways.

@@

\subsection{Mutation and Crossover}
\begin{frame}
  \frametitle{Mutation}
Mutation: random bits in an individual, or possible solution, are randomly changed. Mutation takes bits of rule and changes them to form slightly different rule.  
\begin{figure}
\begin{small}
	\begin{tabular}{|cccc|} \hline
	010 & 011 & 100 & \textbf{101}\\
	a=2 & b=3 & c=4 & \textbf{d=5}\\
	\hline\end{tabular}
	$\rightarrow$
	\quad
	\begin{tabular}{|cccc|} \hline
	010 & 011 & 100 & \textbf{111}\\
	a=2 & b=3 & c=4 & \textbf{d=7}\\
	\hline\end{tabular}
\end{small}
\end{figure}
	\includegraphics[width=0.50\textwidth]{../Talk/MutCrossTrapezoids/mut2345.pdf}
	\includegraphics[width=0.55\textwidth]{../Talk/MutCrossTrapezoids/mut2347.pdf}
\end{frame}

\begin{frame}
  \frametitle{Crossover}
  \begin{itemize}
  	\item Crossover: two individuals swap sequences of bits to form two new individuals.
  	\item Crossover takes 2 rules and creates new rules by swapping bits of old rules.
  \end{itemize}
\end{frame}

@@

\begin{frame}
	\frametitle{Fitness function}
	The fitness function in the algorithm is:
	\begin{equation*}
	\frac{\alpha}{A} - \frac{\beta}{B}
	\end{equation*}

@@

	$B$: \# of normal records.
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Experiments and Results]{Experiments and Results}
\subsection{Two Experiments using Only RLD09}
\begin{frame}
	\frametitle{Experiments Using Only RLD09}
	Experiment 1
	\begin{itemize}
		\item Fuzzy GA was used to create DoS and probe detection rules.

		Both rules were then used together in testing process to identify attacks from testing data set.

@@

		Test set: 26,500 records.

		\item If record is classified as either DoS rule or Probe rule, it is classified as attack; else normal.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusions}
	\begin{itemize}
		\item The fuzzy genetic algorithm had a higher detection rate than a decision tree algorithm in most cases.
		\item Fuzzy genetic algorithms are good at detecting unknown attacks.
		\item The use of fuzzy genetic algorithms in intrusion detection is an effective way of detecting attacks - DR in experiments was in the high 90s.
	\end{itemize}
\end{frame}

>>>>

\newcommand{\tab}[1]{\hspace{.2\textwidth}\rlap{#1}}

@@

\begin{frame}
  \frametitle{RLD09}
	\begin{itemize}
		\item RLD09: created because KDD99 was 10 years old,

@@

\begin{frame}
  \frametitle{Rules}
	\begin{itemize}
		\item A commonly used approach for detecting intrusions is to use rules.
		\item If-Then format: If (\emph{condition}) then (\emph{consequence}).
		\begin{itemize}
			\item Condition: one or more features
			\item Consequence: says if it is an intrusion or not.
			\linespace
			\item If \emph{duration $\leq$ 4} then \emph{intrusion}.
		\end{itemize}				
	\end{itemize}
\end{frame}

@@

\subsection{Determining the Accuracy of an Algorithm}
\begin{frame}
  \frametitle{Determining the Accuracy of an Algorithm}
\begin{table}
Predicted
\begin{tabular}{l|ll}
Actual   & Not Attack & Attack \\ \hline
Not Attack & True Negative (TN) & False Positive (FP) \\
Attack & False Negative (FN) & True Positive (TP)  \\
\end{tabular}
\end{table}
\linespace
\linespace
\begin{center}
Detection rate (DR): percentage of normal and attack activity correctly classified from the total number of data records.
\end{center}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Fuzzy Classification]{Fuzzy Classification}
\subsection{Fuzzy Logic}
\begin{frame}
	\frametitle{Fuzzy Logic}
	\begin{itemize}
		\item Fuzzy logic: used in intrusion detection systems to find degree of certainty of record being attack.
			\begin{itemize}
				\item If it's not clear if activity is an attack, fuzzy logic says where on a spectrum it is and how certain it is of being an attack.
			\end{itemize}
		\linespace
		\linespace
		\linespace
		\item Fuzzy logic rules: similar to rules described before, except that
consequence is certainty factor. 
		\begin{itemize}
			\item If (\emph{duration} = 6.2) then (\emph{the degree of certainty of the record being an attack is 0.8}).
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\begin{figure}
\begin{small}
\begin{tabular}{|cccc|c|cccc|c|} \hline
010 & 011 & 100 & 101   & ... & 001 & 100 & 101 & 111   & Attack\\
a=2 & b=3 & c=4 & d=5   & ... & a=1 & b=4 & c=5 & d=7   &\\  
    &     & Block 1&    &        &     & Block 12& &       & Type\\
\hline\end{tabular}
\caption{Based on [Jongsuebsuk \emph{et al.}, 2013]}
\end{small}
\end{figure}
\end{frame}

\begin{frame}
	\frametitle{Encoding of Features and Rules}
\begin{figure}
\begin{small}
\begin{tabular}{|cccc|c|cccc|c|} \hline
010 & 011 & 100 & 101   & ... & 001 & 100 & 101 & 111   & Attack\\
a=2 & b=3 & c=4 & d=5   & ... & a=1 & b=4 & c=5 & d=7   &\\ 
    &     & Block 1&    &        &     & Block 12& &       & Type\\
\hline\end{tabular}
\end{small}
\end{figure}
\tab{$\downarrow$} \tab{} \tab{} \tab{$\downarrow$}
  \includegraphics[width=0.50\textwidth]{../Talk/MutCrossTrapezoids/mut2345.pdf}
  \includegraphics[width=0.58\textwidth]{../Talk/MutCrossTrapezoids/cross1457.pdf}

Degree of certainty is computed for each of 12 blocks, if sum of those is greater than a threshold, declared as attack.
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Genetic Algorithms]{Genetic Algorithms}
\subsection{GA Overview}
\begin{frame}
  \frametitle{Genetic Algorithms}
	\begin{itemize}
		\item GAs: search technique used to find solutions to problems.
        \item Possible solutions to problems: represented in a variety of problem dependent ways.

@@

\subsection{Mutation and Crossover}
\begin{frame}
  \frametitle{Mutation}
Mutation: random bits in an individual, or possible solution, are randomly changed. Mutation takes bits of rule and changes them to form slightly different rule.

\begin{figure}
\begin{small}
	\begin{tabular}{|cccc|} \hline
	010 & 011 & 100 & \textbf{101}\\
	a=2 & b=3 & c=4 & \textbf{d=5}\\
	\hline\end{tabular}
	$\rightarrow$
	\quad
	\begin{tabular}{|cccc|} \hline
	010 & 011 & 100 & \textbf{111}\\
	a=2 & b=3 & c=4 & \textbf{d=7}\\
	\hline\end{tabular}
\end{small}
\end{figure}

	\includegraphics[width=0.50\textwidth]{../Talk/MutCrossTrapezoids/mut2345.pdf}
	\includegraphics[width=0.55\textwidth]{../Talk/MutCrossTrapezoids/mut2347.pdf}

After mutation: transition from c to d is more gradual.
\end{frame}

\begin{frame}
  \frametitle{Crossover}
  \begin{itemize}
  	\item Crossover: two individuals swap sequences of bits to form two new individuals.
  	\item In IDS: crossover takes 2 rules and creates new rules by swapping bits of old rules.
  \end{itemize}
\end{frame}

@@

\begin{frame}
	\frametitle{Fitness function}
	The fitness function to be maximized is:
	\begin{equation*}
	\frac{\alpha}{A} - \frac{\beta}{B}
	\end{equation*}

@@

	$B$: \# of normal records.

\linespace
\linespace
\linespace
\linespace
Best possible value of $\beta$ is 0.
It's good if $\alpha$ = $A$.

Best possible fitness value is 1.
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Experiments and Results]{Experiments and Results}
\subsection{Two Experiments using Only RLD09}
\begin{frame}
	\frametitle{Experiments Using Only RLD09}
	Experiment 1
	\begin{itemize}
		\item Fuzzy GA was used to create DoS and probe detection rules.

		\item Both rules were then used together in testing process to identify attacks from testing data set.

		\item If record is classified as either DoS rule or Probe rule, it is classified as attack; else normal.

@@

		Test set: 26,500 records.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusions}
	\begin{itemize}
		\item The fuzzy genetic algorithm had a higher detection rate than a decision tree algorithm in most cases.
		\item Fuzzy genetic algorithms are good at detecting unknown attacks.
		\begin{itemize}
			\item Fuzzy GA DR: 99.8\%, Decision Tree DR: 46.7\%
		\end{itemize}

		\item The use of fuzzy genetic algorithms in intrusion detection is an effective way of detecting attacks - DR in all experiments was in the high 90s.
	\end{itemize}
\end{frame}
